---
title: "Friendly Epistaxis"
author: "Tim Radtke"
output: beamer_presentation
---

```{r setup, include=TRUE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Data

```{r}
# load standard packages
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

# load data using readr
efficacy <- read_csv("~/friendly-epistaxis/data/efficacy.csv")
randomization <- read_csv("~/friendly-epistaxis/data/randomization.csv")
subject <- read_csv("~/friendly-epistaxis/data/subject.csv")

all <- subject %>%
  left_join(randomization) %>%
  left_join(efficacy)
```

The currently available data set consists of `r dim(all)[1]` observations and `r dim(all)[2]` variables.

We have `r sum(is.na(all$eye.colour))` NA in the `eye.colour` variable.

We have `r sum(is.na(all$mucus.viscosity))` NA in the `mucus.viscosity` variable.

## Data - Mucus Viscosity

```{r}
all %>%
  ggplot(aes(x = 1, y = mucus.viscosity)) +
  geom_boxplot() +
  theme(text = element_text(size=18))
```

## Data - Mucus Viscosity

```{r}
all %>%
  ggplot(aes(x = 1, y = log(mucus.viscosity))) +
  geom_boxplot() +
  theme(text = element_text(size=18))
```

## Data - Mucus Viscosity

While mucus viscosity is a continuous variable, there are `r sum(all$mucus.viscosity == 0, na.rm = TRUE)` observations that are zero.

Is this a mistake and actually missing data, or correct?

## Data - Mucus Viscosity and Nosebleeds

There does not appear to be an obvious relation between mucus viscosity and nosebleeds requiring hospitalization in the previous year.

```{r, fig.height = 6}
poisson_smooth <- function(...) {
  geom_smooth(method = "glm", 
              method.args = list(family = poisson(link = "log")), ...)
}

all %>%
  ggplot(aes(x = mucus.viscosity, y = previous.year)) +
  geom_jitter(height = 0.1) +
  poisson_smooth() +
  theme_bw() +
  theme(text = element_text(size=18)) +
  labs(subtitle = "Trend line indicates a Poisson GLM smoother.")
```

## Data - Mucus Viscosity and Tissue Use

A more clear relation might exist between mucus viscosity and tissue use:

```{r, fig.height = 6}
binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "binomial"), ...)
}

all %>%
  mutate(tissue.use.bin = ifelse(tissue.use == "MEDIUM", 0, 1)) %>%
  ggplot(aes(x = mucus.viscosity, y = tissue.use.bin)) +
  binomial_smooth() +
  geom_jitter(height = 0.05) +
  theme_bw() +
  theme(text = element_text(size=18)) +
  labs(subtitle = "Trend line indicates a Binomial GLM smoother.")
```

## Data - Nosebleeds Hospitalization and Tissue Use

A similar relationship exists between nosebleeds hospitalizations in the previous year, and tissue use.

```{r}
all %>%
  mutate(tissue.use.bin = ifelse(tissue.use == "MEDIUM", 0, 1)) %>%
  ggplot(aes(x = previous.year, y = tissue.use.bin)) +
  binomial_smooth() +
  geom_jitter(height = 0.05) +
  theme_bw() +
  theme(text = element_text(size=18)) +
  labs(subtitle = "Trend line indicates a Binomial GLM smoother.")
```

## Data - Mucus Viscosity, Hospitalizations, Tissue Use

Given the observed relationships above, one has to be clear about where causal effects have a theoretical foundation. E.g., does tissue use capture the severity of the nosebleed, which in turn is impacted by the mucus viscosity?

If hospitalizations occur more often in cases of large severity (high tissue use), should then the relationship between hospitalizations and mucus viscosity not be larger?

Be wary of adding tissue use as confounder to a model.

## Data - Eye Colour as Control

Estimating reasonable effects on tissue use as done above seems to be difficult though, given that a simple binomial GLM without penalization/prior shows differences across eye colours:

```{r}
all_glm <- all %>%
  mutate(tissue.use.bin = ifelse(tissue.use == "MEDIUM", 0, 1))
fit <- glm(tissue.use.bin~-1+eye.colour, family = binomial(), data = all_glm)
round(summary(fit)$coefficients,2)
round(rethinking::logistic(summary(fit)$coefficients[,1]),2)
```

## Data - Nosebleeds and Hospitalizations

Interestingly, all subjects have at least 2 hospitalizations due to nosebleeds in the previous year (and thus should have had at least 2 nosebleeds).

But even the subjects which have been observed for a year, show mostly less than 2 nosebleeds even for the Placebo.

After filtering for a duration longer than 360 days, we have:

```{r}
all %>%
  filter(duration > 360) %>%
  group_by(arm) %>%
  summarize(mean_nosebleeds = mean(nosebleeds), 
            q85_nosebleeds = quantile(nosebleeds, 0.85),
            min_hosp = min(previous.year)) %>%
  knitr::kable()
```

## Model

### What do we model?

Let $Y_i$ be the number of nosebleeds of patient $i$ and $d$ the number of days for which we have observed her.

Then the rate of nosebleeds per day is given as

$$
\lambda_i = \frac{Y_i}{d_i}, \text{ with } \lambda_i \geq 0
$$

## Model

We are interested in how $\lambda_i$ is affected by the treatment $sdp_i\in\{0,1\}$.

Thus we want to estimate the following model:

$$
\log(\lambda_i) = \beta_0 + \beta_1 \cdot sdp_i
$$

## Model

We hypothesize that the effect of the treatment might vary depending on the mucus viscosity $mv_i$ of patient $i$. One might propose a limited impact after a certain point:

$$
\beta_1 = \gamma_0 + \gamma_1 \cdot \log(1+mv_i)
$$

## Model

The number of nosebleeds of a patient is discrete. Thus propose a count distribution, e.g. Poisson, as generative model:

$$
Y_i \sim \text{Pois}(\lambda_i\cdot d_i)
$$

The Poisson in particular comes with a strong assumption $(\text{E}[Y] = \text{Var}[Y])$ that needs to be checked.

## Model

A first model could thus be given by:

$$
Y_i \sim \text{Pois}(\lambda_i\cdot d_i)
$$

$$
\log(\lambda_i) = \beta_0 + (\gamma_0 + \gamma_1 \cdot mv_i) \cdot sdp_i
$$

## Possible Results

```{r plot_treatment_effect}
load(file = "~/friendly-epistaxis/model/20180704_m_treatment_mucus_plot_objects.rda")

library(ggplot2)
ggplot() +
  geom_jitter(aes(y = nosebleeds/duration*365, x = mucus_viscosity,
                  color = Group), 
              data = all_stan_plot, height = 0.1, alpha = 0.5, size = 1.5) +
  geom_ribbon(aes(x = mucus_viscosity,
                  ymin = lb, ymax = ub, 
                  fill = Group, group = Group), 
              data = effect_estimate,
              alpha = 0.3) +
  geom_line(aes(x = mucus_viscosity, y = mean, color = Group, group = Group),
            data = effect_estimate, size = 1.8) +
  coord_cartesian(ylim = c(-0.1,2.1)) +
  theme_bw() +
  theme(legend.position = "bottom",
        text = element_text(size=18)) +
  labs(x = "Mucus Viscosity", y = "Expected Yearly Nosebleed Rate",
       subtitle = "Ribbon shows 7%-93% credible interval of yearly nosebleed rate.",
       title = "Estimated Effect of Superdupripine")
```